Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-10 sftrends_manu.Rnw
trnd figures in blue and per chng per year

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.

@@

Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.  Seasonal Kendall tests weere also used to describe trends in the model predictions for the observed data.  These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.  Functions in the EnvStats package in R were used for the seasonal Kendall tests \cite{Millard13}.

@@

Formal trend analyses of the observed data suggested that

@@

@@

# ggplot theme
mythm <- theme_minimal() +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

# year only
toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
p1
dev.off()

pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
p2
dev.off()

pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
p3
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

<<dinc10, fig.height = 9, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.  Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} at different conditional quantiles ($\\tau$ = 0.1, 0.5, 0.9).  The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.  Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.  Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend. Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.'>>=

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$). Trends are shown separately for different seasonal groupings from 1976-1995. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp2}   
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$). Trends are shown separately for different seasonal groupings from 1996-2013. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp3}   
\end{figure}

>>>>

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of WRTDS models in tidal waters \cite{Beck15}.

@@

Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.  Seasonal Kendall tests were also used to describe trends in the model predictions for the observed data.  These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.  Functions in the EnvStats package in R were used for the seasonal Kendall tests \cite{Millard13}.

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends from 1976-1995 and 1996-2013 varied considerably between sites and analytes.  Overall,

@@

# color gradients
blues <- RColorBrewer::brewer.pal(9, 'Blues')
collo <- blues[5] #'grey60'
colhi <- blues[9] #'black'

@@

# ggplot theme
mythm <- theme_minimal() +
  theme(
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    legend.position = 'none'
  )

# stupid legend
toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
pleg <- ggplot(toplo1, aes(x = perchg, y = Site_Code)) + 
  geom_point(aes(pch = trnd), colour = blues[7], size = 4, alpha = 0.8) + 
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = collo, high = colhi) +
  mythm + 
  theme(legend.position = 'top')
pleg <- g_legend(pleg)

# year only
p1 <- ggplot(toplo1, aes(x = perchg, y = Site_Code)) + 
  geom_vline(xintercept = 0, linetype = 'dashed') +
  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
    ) + 
  mythm + 
  xlab(expression('% chg yr' ^-1)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
p2 <- ggplot(toplo2, aes(x = perchg, y = Site_Code)) + 
  geom_vline(xintercept = 0, linetype = 'dashed') +
  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
    ) +
  mythm + 
  xlab(expression('% chg yr' ^-1)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
p3 <- ggplot(toplo3, aes(x = perchg, y = Site_Code)) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  geom_point(aes(pch = trnd, colour = abs(perchg)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed, scales = 'free_x'
    ) + 
  mythm + 
  xlab(expression('% chg yr' ^-1)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = collo, high = colhi, guide = F) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

pdf('figs/trndcomp1.pdf', height = 5, width = 9, family = 'serif')
grid.arrange(pleg, p1, ncol = 1, heights = c(0.1, 1))
dev.off()

pdf('figs/trndcomp2.pdf', height = 8, width = 9, family = 'serif')
grid.arrange(pleg, p2, ncol = 1, heights = c(0.06, 1))
dev.off()

pdf('figs/trndcomp3.pdf', height = 8, width = 9, family = 'serif')
grid.arrange(pleg, p3, ncol = 1, heights = c(0.06, 1))
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

<<dinc10, fig.height = 9, fig.width = 7, cache = T, out.width = '0.75\\textwidth', fig.cap = 'Time series of \\ac{din} and flow at station C10.  Subfigure (a) shows the observed \\ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \\ac{wrtds} for the conditional mean response.  The points in subfigure (b) are predictions of observed \\ac{din} and the lines are flow-normalized predictions.  Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.  Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend. Subfigure (e) shows the modelled relationships between \\ac{din}, flow, and time.'>>=

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1976-1995. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp2}   
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1996-2013. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp3}   
\end{figure}

### 2017-01-10 sftrends_manu.tex
trnd figures in blue and per chng per year

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are the estimated direction and magnitude of the trend ($\tau$, varying from -1 to 1 with significance at $\alpha = 0.05$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption{Time series of \ac{din} and flow at station C10.  Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} at different conditional quantiles ($\tau$ = 0.1, 0.5, 0.9).  The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.  Subfigure (c) shows the difference between the model predictions and flow-normalized predictions at the fiftieth conditional quantile.  Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend. Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.}\label{fig:dinc10}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption{Time series of \ac{din} and flow at station C10.  Subfigure (a) shows the observed \ac{din} time series and subfigure (b) shows the annual (water year starting in October) predictions from \ac{wrtds} for the conditional mean response.  The points in subfigure (b) are predictions of observed \ac{din} and the lines are flow-normalized predictions.  Subfigure (c) shows the difference between the model predictions and flow-normalized predictions.  Subfigure (d) shows the flow time series of the San Joaquin River with a locally-estimated (loess) smooth to emphasize the long-term trend. Subfigure (e) shows the modelled relationships between \ac{din}, flow, and time.}\label{fig:dinc10}
\end{figure}
